---
title: "Introdution to R - class 3" 
author: "**Tomek Gaczorek**; tomek.gaczorek@gmail.com"
date: "18.11.22"
output: github_document
---

```{r echo = FALSE}
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, x)
  } else x
}
my_logical <- c(T,T,T,F,F)
```

<!-- `r colorize("**!!! Decription of exercisesis is in green !!!**","green")` -->

<!-- `r colorize("**!!! Advanced parts are marked with the red asterixis !!!**","red")` -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Logical expressions

R as a whole computer architecture (including programming languages) is based on the binary information, the current either flows (TRUE) or not (FALSE, capital letters on purpose) depending on the conditions. These logical values form a distinct type of data and can be, similarly to other types, combined into vectors.

`r colorize("**Ex 1. Create a logical vector with 3 TRUE and 2 FALSE values and call it *my_logical*. Note that they are internal R symbols so you should not use quotation marks.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
my_logical
```

**An advice**  
To make your code shorter you can use T instead of TRUE and F instead of FALSE.

Formally, logical values correspond to 0 (for FALSE) and 1 (for TRUE) and behave like them in every mathematical operations.

`r colorize("**Ex 2. Calculate the sum of *my_logical*.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
sum(my_logical)
```

However, logical vectors have one important distinctive characteristic: they can be used for subsetting. To achieve this you need to provide logical vector with TRUE for the elements you want to keep and FALSE for for the elements you want to discard. The length of the logical vector used for subsetting has to be equal to the number of elements (e.g., columns) in the object we want to subset. 

`r colorize("**Ex 3. Save 6 first rows of built-in *CO2* dataset to a chosen variable. Then, using logical vectors return its 1^st^ and 5^th^ collumn.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
CO2[1:6,c(1,5)]
```

### 2. Comparisons

Normally, no one creates logical vectors on his/her own. There are created automatically as the effect of different comparisons. The most common is *equality*. It is done with double equal sign (*==*).

`r colorize("**Ex 4. Check whether in R: 5 equals 5.00 and whether $\\pi$ equals 3.14.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
5 == 5.00
pi == 3.14
```

The same stands for all others comparisons but the symbols are different:

* **==** - equal
* **!=** - unequal
* **>** - higher
* **>=** - higher or equal
* **<** - lower
* **<=** - lower or equal

**Curiosity**  
Double equal sign is used for equality because single sign is already in use. It serves as an alternative for the assigning arrow, however, for code purity the arrow is the one recommended.  

Note that while comparing two vectors with the symbols shown above, R does not consider the action as one comparison. It rather compares them element by element, recycling shorter vector and resulting in logical vector of longer one's length. It is the same rule as for matematical operations on vectors.

`r colorize("**Ex 5. Manually create two vectors. One of the prime numbers and a second of even numbers. Both should belong to the range <0,10>. Check if they are equal.**","green")`  
Desired outcome:  
```{r echo = FALSE,warning = F, comment= ""}
c(2,3,5,7) == c(2,4,6,8,10)
```

To check whether vectors (as a whole) are identical use *identical()* function.

`r colorize("**Ex 6. Create two integer vectors from 1 to 10. Call them differently and compare them with *identical()* function. Then, change one value within first vector and repeat comparison.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
identical(c(1:10),c(1:10))
identical(c(1:10),c(1:9))
```

The other useful tool is the *%in%* operator. It provides information whether elements of first vector are present in the second one. Note that it is focused on the first vector only, so there is no recycling. 

`r colorize("**Ex 7. Create two character vectors each consisting of a set of individual characters. The first should contain your name and the second one the name of other person from the group. Check how many letter your names have in common. Then, change the order of names and repeat the comparison.**","green")`  

### 3. Exclamation mark

Exclamation mark (*!*) works in R as the symbol of negation. Any logical vector preceded by *!* will result in its reversal - FALSE changed into TRUE and vice versa.

`r colorize("**Ex 8. Having vector of 3 TRUE values, return its negation.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
!c(T,T,T)
```

More typically *!* is used to negate comparisons. Note that the idea is the same: you negate logical vector by negating action which produces it. Remember that negated comparison should be enclosed in parentheses e.g. *!(2 == 2)*.

`r colorize("**Ex 9. Create a sequence of integers from 1 to 100 in which each subsequent element is larger by 3 than the previous one. Then, create logical vector indicating which elements are larger than 50. Do not use '*>*' sign.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
seq(1,100,3)>50
```

### 4. Logical operators

The real power of logic in programming is provided by combining comparisons (use parentheses for clarity). There are two basic operators:

* **&** - and - condition is TRUE if both comparison are TRUE
* **|** - or - condition is TRUE if at least one comparison is TRUE

`r colorize("**Ex 10. For an integer vector from 1 to 10, create logical vector indicating which element is smaller or larger than 5.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
c(1:10)<5 | c(1:10)>5
```

`r colorize("**Ex 11. For an integer vector from 1 to 10, create logical vector indicating which element is divisible by both 2 and 3.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
c(1:10)%%2 == 0 & c(1:10)%%3 == 0
```

**Curiosity**  
R also uses double version of *&* and *|* operators. Their outcome is exactly the same and their role is just related to optimalization. While using double operators, R checks first condition and if not needed, ignore the second one (e.g. in the 'AND' statement if the first condition is FALSE, there is no need to check the second one as the result will always be FALSE). Note that they are often used when conditions are severly time-consuming.  

### 5. *which()* function

Frequently, the question is not about logical vectors themselves but rather about which element of a vector fulfills given condition. The answer is provided by *which()* function. It takes comparison as an argument and return a vector of indexes.

`r colorize("**Ex 12. Construct a vector with first 20 integers divisible by 3. Which elements of it are larger or equal 21? **","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
which(seq(3,3*20,3)>=21)
```

`r colorize("**Ex 13. Having indexes from the previous exercise, return the values from the vector.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
seq(3,3*20,3)[which(seq(3,3*20,3)>=21)]
```

### 6. Subsetting with the logical expressions

As stated before subsetting can be done directly with logical vectors (TRUE for each kept element) or indexes generated by *which()* function. In practice, it is even simpler. All you need to provide is a condition instead of a coordinate e.g. *vector[condition]* will return only the elements fulfilling given condition.  
The logic behind is as follows: condition generates a logical vector >> coordinates are denoted as a series of TRUEs and FALSEs elements in this logical vector >> returned are elements of the vector you want to subset, for which coordinate equals TRUE.  

`r colorize("**Ex 14. For an integer vector from 1 to 100, return elements higher than vector's mean.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
c(1:100)[c(1:100)>mean(c(1:100))]
```

The same pattern applies for subsetting data frames. 

`r colorize("**Ex 15. Using built-in dataset *CO2*, return observation for *Qn2* plant.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
CO2[CO2$Plant == "Qn2",]
```

**An advice**  
Note that to obtain the logical vector with one element for each row, you need to make comparison based on a collumn e.g. *my_data[my_data$my_column != 5,]* would result in the observations (including all columns) from *my_data* in which the value for *my_collumn* does not equal 5.   

`r colorize("**Ex 16. Using built-in dataset *CO2*, return observation from Mississippi chilled plant with an uptake higher than 20 ummol/m^2^ x s.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
CO2[CO2$Type == "Mississippi" & CO2$Treatment == "chilled" & CO2$uptake > 20,]
```

### 7. IF, ELSE statements

The next important application of conditions is decision making. It is extremely useful when you want your code to behave differently depending on conditions. It is done with *if()* and *else()* statements with the following syntax (note curly brackets):  
```{r echo = T,eval=F, comment= ""}
if(your_condition){
  # do something if condition is TRUE
} else {
  # do something otherwise
}
```

To perform it properly be aware of 3 things:

* your condition needs to result in a single TRUE or FALSE (not a vector with 2 or more elements). Think about it as real-life decision - yes or no.
* *else()* statement makes sense only when preceded by *if()* statement.
* *else()* statement is optional. If not used, R simply does not perform any action when condition is FALSE

`r colorize("**Ex 17. Assume x is a real number. Write piece of code printing positive number when x is negative and vice versa. Check it for x equals -1 and 2.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
1
-2
```

**An advice**  
You can print a value by simply typing it without any function or with the use of *print()*.

Sometimes two options are not enough. You can introduce more conditions with *else if()*

```{r echo = T,eval=F, comment= ""}
if(condition_1){
  # do something if condition_1 is TRUE
} else if(condition_2){
  # do something if condition_1 if FALSE but condition_2 is TRUE
} else {
  # do something otherwise
}
```

`r colorize("**Ex 18. Assume x is a real number. Write piece of code printing 'P' if the number is positive, 'N' if negative and 'Z' if it equals 0. Check it for -1,0 and 2.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
"N"
"Z"
"P"
```

`r colorize("******************ADVANCED********************","red")`  
Making a proper choice often require making subsequent decisions based on previous choices. To do this, nest one *if()* statement into another (there is no limit of steps).
```{r echo = T,eval=F, comment= ""}
if(condition_1){
  # do something if condition_1 is TRUE
  if(condition_2){
    # do something if both conditions are TRUE
  } else {
    # do something if condition_1 is TRUE but condiiton_2 if FALSE
  }
} else {
  # do something if condition_1 is FALSE
}
```

**Curiosity**  
1. Note that R Studio by default intends blocks of code to make reading easier.  
2. Note that nested *if()* statement behave in the same way as single one with conditions seperated with *&* operator.

`r colorize("**Ex 19. Assume n is one of the DNA nucleotides. Write piece of code detecting two things: (1) whether it belongs to [purines or pyrimidines](https://library.med.utah.edu/NetBiochem/pupyr/pp.htm) and (2) what is its complementary nucleotide. Check it for all nucleotides (A,T,C,G).**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
"Purine"
"T"
"Pyrimidine"
"A"
"Pyrimidine"
"G"
"Purine"
"C"
```

`r colorize("**Ex 20. Copy and modify previous code to obtain, as a result, the vector consisting: checked nucleotide, type (purines or pyrimidines) and complementary nucliotide.**","green")`   
Desired outcome:  
```{r echo = FALSE, comment= ""}
c("A","Purine","T")
c("T","Pyrimidine","A")
c("C","Pyrimidine","G")
c("G","Purine","C")
```

`r colorize("************************************************************","red")` 

### 8. Loops

Until now, you should know how to force your code to behave differently based on conditions. Imagine, however, checking thousands or millions of values. It does not make sense to do it manually, right? The proper tool for it is a loop.

The basic one is the *FOR* loop. It executes given lines of code independently for each element of a given vector, one by one. The syntax is as follow:
```{r echo = T,eval=F, comment= ""}
for(iterator in your_vector){
  # do something for each element (iterator) of your_vector
  # e.g. 'iterator+2' will print each element of your vector enlarged by 2 
}
```
**An advice**    
1. Inside a loop, given element of a vector is represented by an *iterator* (letter *i* is often used by convention).  
2. Due to the fact that *FOR* loop is often used to iterate over range of numbers, R let you to ommit *c()* function while creating such vectors. Then, the syntax is e.g. *for(iterator in 1:1000){...*.

`r colorize("**Ex 21. Create *FOR* loop that will print all numbers from 1 to 2000.**","green")`

Unless saved, outcome obtained within a given loop cycle is overwritten when iterator is replaced with the next value. What is received at the end is just the result for last element of a vector. To save each outcome, create an empty vector (just *c()*) and save there your outcome during each loop cycle. Remember that: 

1. Empty vector needs to be created before the loop. Note that creation of a new vector within a loop will result in erasing previously saved values.  
2. Outcomes can be saved in three ways:  
* you can append a vector by combining previous and new value e.g. *outcomes <- c(outcomes,new_value)*. For the large number of values (e.g. millions), however, it becomes extremely computationally demanding (rewritting variable).     
* if your iterator represents subsequent numbers starting from 1, you can assign new value to the specific position in a vector, indicated by the iterator e.g. *outcomes[iterator] <- new_value*. Note that if iterators start from 2 you can use *iterator-1* instead.  
* you can create a variable that will count loop cycles and serve as position indication. To do this, create a variable which equals 0 and then add one to it with subsequent cycles. Then, use it as *outcome[counter] <- new_value*. 

`r colorize("**Ex 22. Using *FOR* loop, return modified vector from 1 to 1000 where every second value will be reduced by 100.**","green")`  
First 100 elements of desired vector:  
```{r echo = FALSE, comment= ""}
bee <- c()
for(i in 1:1000){
  if(i %% 2 == 0){
    bee[i] <- i-100
  } else {
    bee[i] <- i
  }
}
bee[1:100]
```

`r colorize("******************ADVANCED********************","red")`   
The second loop is called *WHILE*. The idea is similar but *WHILE* cycle as long as the given condition is TRUE. The syntax is as follow:
```{r echo = T,eval=F, comment= ""}
while(your_condition){
  # do something as long as your_condition is TRUE
}
```

*WHILE* loop is a little bit dangerous as improperly written can last forever. There are 2 main techniques used:  
1. In the condition, place some variable that will be changed in each loop cycle. After desired number of cycles, it should reach a threshold which falsifies the condition e.g. consequent number addition.  
2. you can use *checker* variable which eqauls TRUE. It needs to be created before the loop. The initial condition should be *(checker == TRUE)*. Then inside the loop you should include *IF* statement changing *checker* to FALSE if some other condition is fulfilled. 

`r colorize("**Ex 23. Using WHILE loop create a vector with [Fibonacci sequence](https://en.wikipedia.org/wiki/Fibonacci_number) consisting numbers lower than 1000.**","green")`  
Desired outcome:  
```{r echo = FALSE, comment= ""}
out <- c(0,1)
while(out[length(out)] < 1000){
  out[length(out)+1] <- out[length(out)]+out[length(out)-1]
}
out[-length(out)]
```

**An advice**  
Last element of a vector can be obtained by placing its length as a coordinate.

`r colorize("************************************************************","red")` 

### 9. Next classes

**Please, send me the script from today classes.**

During the next class you will learn about function from the *apply* family, writting your own functions and plotting.

### 10. Homework

**Prepare your homework in a form of script file (.R) and include your surname in its name.**

Include all subsequent steps in a script file.   
1. Create a vector from 1 to 50. Then, using R only, calculate how many elements of this vector are higher than the positive square root of 456.    
2. Using built-in *mtcars* dataset, return (in one table) observations with the lowest and the highest value for the gross horsepower. Don't look for the rows' indexes manually.  
3. Assume x is an integer. Create an *'if'* statement printing word "divisible" if x is divisible by 5 and "undivisible" if it is not divisible by 5.  
4. Using *FOR* loop, create 100, 8-letters "words" created from random letters drawn without replacement. All 8 letters should be paste together. Of course, the "words" don't need to mean anything.    
5. Modify the loop above such as all words contain at least one vowel.  
`r colorize("******************ADVANCED********************","red")`   
6. Imagine you are a gambler and you play a game that randomly draws (with replacement) 3 letters from the alphabet. If 3 identical letter are drawn you win 100€ and nothing otherwise. Each game costs 1€ and you start with 200€. Play until you run out of money. Remember that won money gets back to pool. Print how many times you played.  
`r colorize("************************************************************","red")` 

**An advice**   
1. Positive square root can be calculated with *sqrt()* function  
1. Alphabet can be obtained from built-in vector *LETTERS*.  
2. For sampling use *sample()* function.